\bf Comparing the Power of Plot Designs to Reveal Correlation

### 2016-01-12 paper/paper.Rnw
Cleanup

\documentclass[12pt]{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage{wrapfig,float}
\usepackage[font=small,skip=5pt]{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage[section]{placeins}
\usepackage{afterpage}

@@

<<setup, echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
library(lme4)
library(ggplot2)
library(mnormt)
library(tidyr)
library(knitr)
# Convert reshape2 to tidyr, melt to gather - done
#library(reshape2)
library(nullabor)
library(dplyr)
library(mFilter)
library(grid)
library(gridExtra)
library(stringr)
#library(ipapi)
library(maps)
library(moments)
#library(pander)
#library(splitstackshape) - changed cSplit to unnest
#library(readr)
#library(lubridate)
# Convert these to read_csv at some point
correl <- read.csv("correlations.csv", stringsAsFactors = FALSE)
full <- read.csv("full.csv", row.names=1, stringsAsFactors = FALSE) 
prediction <- read.csv("sas/predfit.csv") #Large file, very slow
time.fit <- read.csv("sas/time.csv")
powercoef <- read.csv("sas/PowerEstimates.csv")

colnames(prediction) <- c("rabs", "t", "alpha", "rsgn", "smoothed", "design", "subj_id", "weights", "detected", "log_time", "RandomEf", "FixedEf", "residual")
prediction <- mutate(prediction, r = ifelse(rsgn==1, rabs, -rabs), design = ifelse(design==0, "Overlaid Lines", "Scatterplot"), design = ifelse(smoothed==1, paste(design, "(Smoothed)", sep=" "), design),  diff = round(RandomEf - FixedEf,5))

full$subj_id <- factor(full$ip_address, labels=1:207)
y <- data.frame(table(full$subj_id))
keep <- y$Var1[y$Freq>9]

full <- mutate(full, design = ifelse(test_param ==1, "Overlaid Lines", "Scatterplot"),
               r = round(correlation, 1),
               rsgn = ifelse(r >= 0, 1, 0),
               rabs=abs(r),
               alpha = log(0.05/0.95),
               attempts = sapply(strsplit(as.character(full$response_0), ","), length), #count number of selections in response_0 column
               weights = 1/attempts)
full <- full %>% transform(response_0= strsplit(as.character(response_0), ",")) %>% unnest(response_0)
#split multiple selections in response_0 into different rows
full <- mutate(full, correct = ifelse(response_0 == actual_location, 1, 0),
                   score = ifelse(correct==1, (20-attempts)/19, 0))
full <- select(full, subj_id, correct, score, r, rsgn, rabs, smoothed, n, design, log_time, response_0, actual_location, pic_id, choice_reason, attempts, weights, alpha, conf_level)

full.sub <- filter(full, subj_id %in% keep) #keeping only >9 selections in actual modelling to better estimate random effects

bylineup <- summarise(group_by(full.sub, pic_id, n, smoothed, design, r, actual_location), total=sum(score), correct=sum(correct), evals=sum(attempts*weights))
bylineup <- mutate(bylineup, percent = total/evals)
bylineup.sub <- summarise(group_by(full.sub, pic_id), total=sum(score), evals=sum(weights*attempts))

gen_true_data <- function(n, r, dep=TRUE, smoothed = FALSE) {
  d <- data.frame(rmnorm(n, c(0, 0), matrix(c(1, r, r, 1), 2, 2))) 
  if (dep) {
    d$X1 <- as.vector(arima.sim(list(ar=n/100), n, innov=d$X1))
    d$X2 <- as.vector(arima.sim(list(ar=n/100), n, innov=d$X2))
  }
  if (smoothed) {
    d$X1 <- as.vector(hpfilter(d$X1, freq=1, type="lambda", drift = FALSE)[[2]])
    d$X2 <- as.vector(hpfilter(d$X2, freq=1, type="lambda", drift = FALSE)[[2]])
    d <- as.data.frame(d)
  }
  d$X1 <- scale(d$X1)
  d$X2 <- scale(d$X2)
  d$t <- 1:n
  return(d)
}

gen_null <- function(n, m=20, dep=TRUE, smoothed = FALSE){
  nd <- NULL
  for(i in 1:(m-1)) { 
    d <- data.frame(rmnorm(n, c(0, 0), matrix(c(1, 0, 0, 1), 2, 2)))
    if (dep) {
      d$X1 <- as.vector(arima.sim(list(ar=n/100), n))
      d$X2 <- as.vector(arima.sim(list(ar=n/100), n))
    }
    if (smoothed) {
      d$X1 <- as.vector(hpfilter(d$X1, freq=1, type="lambda", drift = FALSE)[[2]])
      d$X2 <- as.vector(hpfilter(d$X2, freq=1, type="lambda", drift = FALSE)[[2]])
      d <- as.data.frame(d)
    }
    d$X1 <- scale(d$X1)
    d$X2 <- scale(d$X2)
    nd <- rbind(nd, d)
  } 
  nd$.n <- rep(1:(m-1), each = n) 
  nd$t <- 1:n
  return(nd)
}
@

@@

Where $\hat\pi_{i} = (x_{i}+1)/(n_{i}+1)$, where $x$ is the number of times a true lineup was correctly identified and $n$ is the Welch-Satterwaite estimate for the degrees of freedom.

@@

The variation in the covariance parameter allows for control in the correlation in the pairs of data. The null data has no structure, so uncorrelated pairs are produced by setting $\rho = 0$. True data pairs are generated with $\rho = \pm \{0.3, 0.5, 0.7, 0.9\}$. Pilot testing found that the rate of picking the true data plot fell to approximately $5\%$, the $1/m$ rate of randomly picking the true plot, for pairs with correlation below $\rho = 0.3$. Further to this, relationships weaker than this point are often of little interest to analysts. Hence we found it unnecessary to test a more full range of correlations. Simulating data does not produce correlations exactly equal to $\rho$, so simulated data pairs were accepted as real data if pairwise correlation between the two series is within 0.015 of the desired value of $\rho$. The null pairs often had correlation generated spuriously through the simulation process. This was particularly problematic for generating small sample data, but setting $\beta = max(0.5,100t)$ for both true and null data reduced spurious correlation to manageable levels.
For particularly large time series, the AR(1) model can create particularly 'jagged' lines that may be unrealistic for many actual applications. To counter this, additional sets of data were created with a Hodrick-Prescott filter to capture the trend of the model and remove the noise in the cycle component. This filter creates a trend by penalising both errors in smoothness and in fit. The trend of the series, $\tau_{t}$ is defined by the equation:

@@

Subjects were allowed to make multiple selections, which is handled by probability weighting in the model. This increases our data from 1684 to 1849 evaluations.

@@

\iffalse
\begin{figure}[htbp]
  \centering
<<powerdiff, dependson='setup', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
#prop <- bylineup[order(bylineup$r, bylineup$n, bylineup$smoothed),c("smoothed", "n", "evals", "total", "r", #"design")]
#prop <- summarise(group_by(prop, smoothed, n, r), l.attempt=sum(ifelse(design=="Overlaid Lines", total, 0)),
#                                                  l.correct=sum(ifelse(design=="Overlaid Lines", score, 0)),
#                                                  s.attempt=sum(ifelse(design=="Scatterplot", total, 0)),
#                                                  s.correct=sum(ifelse(design=="Scatterplot", score, 0)))
#prop <- mutate(prop, l.adj = (l.correct+1)/(l.attempt+1), s.adj = (s.correct+1)/(s.attempt+1),
#               df = s.attempt + l.attempt - 1, difference=s.adj-l.adj, t = qt(0.975, df),
#               upper = difference + t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt),
##               lower = difference - t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt))
#ggplot(data=prop) + geom_errorbar(aes(x=r, ymax=upper, ymin=lower)) + geom_point(aes(x=r, y=difference)) +
#        geom_hline(aes(yint=0, colour="red")) + labs(x="Correlation", y="Scatterplot power minus Overlaid Lines") + facet_grid(n~smoothed)
@
 \caption{The difference in estimated power (scatter plot power minus overlaid line power) to reveal the true plot by smoothness and sample size with a 95\% confidence interval. The red line indicates the point where both plot designs are equally powerful while values above this line indicate that the scatter plot is more powerful than the overlaid line graph. The largest differences appear in negative correlation, where the scatter plot is more powerful.}
 \label{powerdiff}
\end{figure}
\fi

@@

Figure~\ref{samplesize} demonstrates the effects of changing sample size for unsmoothed data. The scatter plot has increased power to detect correlation almost universally for any increase. Sample size influences the overlaid lines differently, they appears to benefit for positively correlated true plots, but it does not improve the weak performance in negative correlation.
Figure~\ref{time} and Figure~\ref{confidence} give a breakdown of the interactions between the response variables: plot detection, time and confidence. Figure~\ref{time} is the histogram of the time taken to record a response, split by plot design and sign of correlation and true plot detection. The time data was highly positively skewed with a maximum of 1760 seconds so the variable was log-transformed. We found that the subject could identify the true plot out of the overlaid line graph 33.6\% of the time and it took on average $e^{3.442}$ = 31.2 seconds to make a choice. Successful detections were slightly faster than unsuccessful decision, taking on average 28.7 seconds to pick the true plot and 32.6 seconds if a null was selected. The scatter plots performed much better, with the true plot picked 52.9\% of the time, with an average of 22.7 seconds required to make a decision. If the true plot was detected it took only 18.6 seconds, and 29.3 seconds for a null plot selection.

@@

There are two major reasons why a subject may fail to detect the true plot out of a lineup. Either the plot design was unable to display correlation for that particular combination of factors (and the analyst would commit a Type II error by not rejecting the null hypothesis), or a null plot that had a greater correlation than the true plot was selected instead (and the analyst would commit a Type I error by rejecting the null hypothesis). We next can analyse these kinds of errors with the pin plots found in Figure~\ref{pins}. The frequency of selection of each plot in each lineup, either true or null, is represented by a pin, placed at the value of the correlation between the two variables shown in that plot. The correlation of the selected plot is on the x-axis and the frequency of selection is plotted on the y axis. The correlation of the simulated (true) data ($\rho = \pm \{0.3, 0.5, 0.7, 0.9\}$ in the plot, the sample size ($t= \{12, 24, 48, 96\}$), and the plot design (unsmoothed/smoothed, scatterplot/lineplot) is used to facet the plots. The blue pins indicate the true plots and the red pins indicate the null plots. If the plot design has the power to display correlation, we would expect the highest correlation plot to be selected, leading to a Type I error where a plot from the null distribution spuriously had high correlation. We find that this is generally not the case. The selections appear to have a wide range of correlations, indicating that if a subject selected a null plot it was typically not due to excess spurious correlation. Instead the failure to select the highest correlated plot indicates that a Type II error was committed rather than Type I as the plot designs did not have the power to differentiate between medium and low correlation plots. However, for the scatter plot, whenever there was a plot with a very strong correlation, usually in the form of the true plot, it was overwhelmingly selected. For the overlaid lines this mostly occurred only for plots with strong positive correlation, which is consistent with its general lack of power to reveal negative relationships in data.

\begin{figure}[htbp]
  \centering
<<pins, dependson='setup', fig.width=6, fig.height=10, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
count.setup <- select(full, pic_id, response_0, actual_location, 
              design, smoothed, n)
count.setup$response_0 <- as.character(count.setup$response_0)
counts <- data.frame(table(count.setup$pic_id, count.setup$response_0))
counts$Var1 <- as.numeric(as.character(counts$Var1))
counts$Var2 <- as.numeric(as.character(counts$Var2))
colnames(counts) <- c("lineup", "plot", "count")
pin <- merge(counts, bylineup, by.x="lineup", by.y="pic_id")
pin <- select(pin, lineup, plot, count, smoothed, actual_location, n, design, r, evals)
pin$r <- factor(pin$r)
pin$base <- 0
pin$Correlation <- diag(as.matrix(correl[pin$lineup, (pin$plot+2)]))
pin <- mutate(pin, true = ifelse(plot==actual_location, "True plot", "Null plot"), percent = count/evals)
pin.a <- ggplot(data=filter(pin, smoothed==0 & design == "Scatterplot")) + geom_errorbar(aes(x=Correlation, ymin=base, ymax=percent, colour=true)) +
       geom_point(aes(x=Correlation, y=percent, colour=true)) + coord_fixed() + facet_grid(n~r) + labs(title="Unsmoothed Scatterplot", y="Percent Chosen") + theme(legend.position="none")
pin.b <- ggplot(data=subset(pin, smoothed==1 & design == "Scatterplot")) + geom_errorbar(aes(x=Correlation, ymin=base, ymax=percent, colour=true)) +
  geom_point(aes(x=Correlation, y=percent, colour=true)) + coord_fixed()  + facet_grid(n~r) + labs(title="Smoothed Scatterplot", y="Percent Chosen") + theme(legend.position="none")
pin.c <- ggplot(data=subset(pin, smoothed==0 & design == "Overlaid Lines")) + geom_errorbar(aes(x=Correlation, ymin=base, ymax=percent, colour=true)) +
  geom_point(aes(x=Correlation, y=percent, colour=true)) + coord_fixed()  + facet_grid(n~r) + labs(title="Unsmoothed Overlaid Lines", y="Percent Chosen") + theme(legend.position="none")
pin.d <- ggplot(data=subset(pin, smoothed==1 & design == "Overlaid Lines")) + geom_errorbar(aes(x=Correlation, ymin=base, ymax=percent, colour=true)) +
  geom_point(aes(x=Correlation, y=percent, colour=true)) + coord_fixed()  + facet_grid(n~r) + labs(title="Smoothed Overlaid Lines", y="Percent Chosen") + theme(legend.position="none")
grid.arrange(pin.a, pin.b, pin.c, pin.d, ncol=1)
@
  \caption{Pin plots of the subject selections of each lineup, by design, smoothness, true plot correlaition and sample size. On the x-axis is the correlation of the plot (there is potentially spurious correlation in the null plots), and on the y-axis is the proportion a plot was chosen. Each plot contains data from the three lineups with the requisite characteristics. Red pins indicate null plots and blue indicates the true plot, occasionally the true plot does not seem to appear as it has a proportion of zero and is visually blocked by the red pins. Strongly correlated plot seem to be selected regardless of if the relationship is real or spurious.}
  \label{pins}
\end{figure}

@@

\subsection{Modelling Subject Responses}

@@

\begin{table}[htp]
\caption{Weighted Mixed Effects Logit Model for the predicted power of plot design. Sample size for scatter plot evaluations: 900, overlaid lines: 949, total: 1849. Default plot design is overlaid lines. The model is fitted using the GLIMMIX procedure in SAS.}
\label{tab1}
\centering
<<powerfit, dependson='setup', results='asis', echo=FALSE, message=FALSE, error=FALSE, warning=FALSE>>=
sasdata <- select(full.sub, rabs, n, alpha, rsgn, smoothed, design, subj_id, weights, correct, log_time)
sasdata$rabs <- as.numeric(sasdata$rabs)
sasdata$subj_id <- as.numeric(sasdata$subj_id)
sasadata <- mutate(sasdata, design = ifelse(design=="Scatterplot", 1, 0))
#write.csv(subsetted, "sas/choices.csv", row.names = FALSE)
powercoef <- powercoef[,-1]
rownames(powercoef) <- c("Intercept", "abs(r)", "sign(r)", "abs(r)*sign(r)", "smoothed", "t", "design", "design*abs(r)", "design*sign(r)", "design*abs(r)*sign(r)", "design*smoothed", "design*t")
colnames(powercoef) <- c("Estimate", "standard error", "df", "t-value", "p-value")
kable(powercoef, format="latex", digits=2)
@
\end{table}

Table \ref{tab1} summarises the model estimates. It can be confusing the interpret the model estimates, and at first glance it appears the negative plot coefficient implies that overlaid lines are substantially more powerful than the scatter plot, but this is misleading and interpreting the full model involves combining the estimates for the different interactions with $\rho$. $\beta_{2}$ and $\beta_{8}$ model the power difference between low and high correlations, while $\beta_{3}, \beta_{4}, beta_{9}$ and $\beta_{10}$ model any increase in power appearing only in positive correlation. 
Taking these into account, the scatter plot shows superior performance relative to the overlaid lines at higher values of abs($\rho$). It needs to be restated that we only tested correlations at  $0.3, 0.5, 0.7$ and $0.9$, so the results below an absolute value of $0.3$ are unreliable and may be artifacts of regression smoothing methods.

%The correlation:sign interaction shows that the scatter plot has close to symmetric performance for positive (correlation:sign + correlation) and negative (-correlation) correlation while the overlaid line graph is substantially weaker for negative values of correlation.
Smoothing out the excess 'noise' from a time series had a minimal effect on scatter plots(...), it did have a significant positive impact on the power of overlaid line graphs as smoothing induces a large visual difference (Figure~\ref{smoothing}. Increasing the sample size, $t$, improved the power of the scatter plot but it was not statistically significant for the overlaid line graph.

>>>>

\documentclass[12pt]{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage{wrapfig,float}
\usepackage[font=small,skip=5pt]{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage[section]{placeins}
\usepackage{afterpage}
\usepackage{bbm}

@@

<<setup, echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
library(lme4)
library(ggplot2)
library(mnormt)
library(tidyr)
library(knitr)
library(nullabor)
library(dplyr)
library(mFilter)
library(grid)
library(gridExtra)
library(stringr)
library(maps)
#library(readr)
#library(lubridate)
# Convert these to read_csv at some point
correl <- read.csv("correlations.csv", stringsAsFactors = FALSE) ##Correlations of every plot in every lineup
full <- read.csv("full.csv", row.names=1, stringsAsFactors = FALSE) ##MTurk evaluations
prediction <- read.csv("sas/predfit.csv") ##Large file, very slow, fitted data from SAS
time.fit <- read.csv("sas/time.csv") ##coefficients for time model from SAS
powercoef <- read.csv("sas/PowerEstimates.csv") ##coefficients for power model from SAS

full$subj_id <- factor(full$ip_address, labels=1:207)
y <- data.frame(table(full$subj_id))
keep <- y$Var1[y$Freq>9]
##Recording who evaluated at least ten lineups. Done prior to splitting multiple selections

full <- full %>% mutate(design = ifelse(test_param ==1, "Overlaid Lines", "Scatterplot"),
                         r = round(correlation, 1),
                         rsgn = ifelse(r >= 0, 1, 0),
                         rabs = abs(r),
                         alpha = log(0.05/0.95),
                         attempts = sapply(strsplit(as.character(full$response_0), ","), length), 
                         ##count number of selections in response_0 column
                         weights = 1/attempts) %>%
  transform(response_0= strsplit(as.character(response_0), ",")) %>% unnest(response_0) %>%
  ##split multiple selections in response_0 into different rows
  mutate(correct = ifelse(response_0 == actual_location, 1, 0), 
         score = ifelse(correct==1, (20-attempts)/19, 0)) %>%
  select(subj_id, correct, score, r, rsgn, rabs, smoothed, n, design, log_time, response_0,
         actual_location, pic_id, choice_reason, attempts, weights, alpha, conf_level)

full.sub <- filter(full, subj_id %in% keep) 
##keeping only subjects with >0 evaluations in actual modelling to better estimate random effects

bylineup <- summarise(group_by(full, pic_id, n, smoothed, design, r, actual_location), total=sum(score), correct=sum(correct), evals=sum(weights))
bylineup <- mutate(bylineup, percent = total/evals)
bylineup.sub <- summarise(group_by(full.sub, pic_id), total=sum(score), evals=sum(weights))
bylineup.sub <- mutate(bylineup.sub, percent = total/evals)

##Summarise data by lineup

#sasdata <- select(full.sub, rabs, n, alpha, rsgn, smoothed, design, subj_id, weights, correct, log_time)
#sasadata <- mutate(sasdata, rabs = as.numeric(rabs),
#                   subj_id = as.numeric(subj_id)
#                   design = ifelse(design=="Scatterplot", 1, 0))
#write.csv(subsetted, "sas/sasdata.csv", row.names = FALSE)
##The code that is exported to fit the model into SAS, see sas/model fit.sas

# data.frame(expand.grid(rabs=rabs, t=t, alpha=log(.05/0.95), rsgn=c(0,1),
#                                      smoothed=c(0,1), design=c(0,1),
#                                      subj_id=full.sub$subj_id))
#pred <- subset(pred, smoothed==0 | smoothed == 1 & t ==48 | smoothed ==1 & t ==96)
#write.csv(pred, "sas/predgrid.csv", row.names=FALSE) # these lines are done, but its a large file so commented out for knitr
#Code exported to SAS to predict new values

colnames(prediction) <- c("rabs", "t", "alpha", "rsgn", "smoothed", "design", "subj_id", "weights", "detected", "log_time", "RandomEf", "FixedEf", "residual")
prediction <- mutate(prediction, r = ifelse(rsgn==1, rabs, -rabs),
                     design = ifelse(design==0, "Overlaid Lines", "Scatterplot"), 
                     design = ifelse(smoothed==1, paste0(design, " (Smoothed)"), design),  
                     diff = round(RandomEf - FixedEf,5))
##Data from SAS

gen_true_data <- function(n, r, smoothed = FALSE) {
  d <- data.frame(rmnorm(n, c(0, 0), matrix(c(1, r, r, 1), 2, 2))) 
  d$X1 <- as.vector(arima.sim(list(ar=n/100), n, innov=d$X1))
  d$X2 <- as.vector(arima.sim(list(ar=n/100), n, innov=d$X2))
  if (smoothed) {
    d$X1 <- as.vector(hpfilter(d$X1, freq=1, type="lambda", drift = FALSE)[[2]])
    d$X2 <- as.vector(hpfilter(d$X2, freq=1, type="lambda", drift = FALSE)[[2]])
    d <- as.data.frame(d)
  }
  d$X1 <- scale(d$X1)
  d$X2 <- scale(d$X2)
  d$t <- 1:n
  return(d)
}
##Generates a pair of data, length n, correlation r

gen_null <- function(n, m=20, smoothed = FALSE){
  nd <- NULL
  for(i in 1:(m-1)) {
    d <- data.frame(X1 = as.vector(arima.sim(list(ar=n/100), n)),
                    X2 = as.vector(arima.sim(list(ar=n/100), n)))
    if (smoothed) {
      d$X1 <- as.vector(hpfilter(d$X1, freq=1, type="lambda", drift = FALSE)[[2]])
      d$X2 <- as.vector(hpfilter(d$X2, freq=1, type="lambda", drift = FALSE)[[2]])
      d <- as.data.frame(d)
    }
    d$X1 <- scale(d$X1)
    d$X2 <- scale(d$X2)
    nd <- rbind(nd, d)
  } 
  nd$.n <- rep(1:(m-1), each = n) 
  nd$t <- 1:n
  return(nd)
}
##Generates m-1 pairs of uncorrelated data, length n
@

@@

Where $\hat\pi_{i} = (x_{i}+1)/(n_{i}+1)$, $x$ is the number of times a true lineup was correctly identified and $n$ is the Welch-Satterwaite estimate for the degrees of freedom.

@@

The variation in the covariance parameter allows for control in the correlation in the pairs of data. The null data has no structure, so uncorrelated pairs are produced by setting $\rho = 0$. True data pairs are generated with $\rho \in \pm \{0.3, 0.5, 0.7, 0.9\}$. Pilot testing found that the rate of picking the true data plot fell to approximately $5\%$, the $1/m$ rate of randomly picking the true plot, for pairs with correlation below $\rho = 0.3$. Further to this, relationships weaker than this point are often of little interest to analysts. Hence we found it unnecessary to test a more full range of correlations. Simulating data does not produce correlations exactly equal to $\rho$, so simulated data pairs were accepted as real data if pairwise correlation between the two series is within 0.015 of the desired value of $\rho$. The null pairs often had correlation generated spuriously through the simulation process. This was particularly problematic for generating small sample data, but setting $\beta = min(0.5,100t)$ reduced spurious correlation to manageable levels.
For particularly large time series, the AR(1) model can create particularly 'jagged' lines that may be unrealistic for many actual applications. To counter this, additional sets of data were created with a Hodrick-Prescott filter to capture the trend of the model and remove the noise in the cycle component. This filter creates a trend by penalising both errors in smoothness and in fit. The trend of the series, $\tau_{t}$ is defined by the equation:

@@

Subjects were allowed to make multiple selections from each lineup evaluated, additional selections on lineup $i$ by subject $j$ are penalised according to a score defined by:
\begin{equation}
\label{score}
{score}_{ij} =\frac{m-s_{ij}}{m-1} \mathbbm{I}_{ij}
\end{equation}

where $\mathbbm{I}_{ij} = 1$ if the true plot is detected, $0$ otherwise, $s_{ij}$ is the number of selections made and $m = 20$ is the number of plots in the lineup.

However, for the model fit to power, multiple selections are seperated and given a weighting equal to to inverse of the number of selections made in that lineup evaluation. This increases our data from 1684 to 1849 evaluations.

@@

\iffalse
\begin{figure}[htbp]
  \centering
<<powerdiff, dependson='setup', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
prop <- select(bylineup, smoothed, n, evals, total, r, design)
prop <- summarise(group_by(prop, smoothed, n, r), l.attempt=sum(ifelse(design=="Overlaid Lines", evals, 0)),
                                                  l.correct=sum(ifelse(design=="Overlaid Lines", total, 0)),
                                                  s.attempt=sum(ifelse(design=="Scatterplot", evals, 0)),
                                                  s.correct=sum(ifelse(design=="Scatterplot", total, 0)))
prop <- mutate(prop, l.adj = (l.correct+1)/(l.attempt+1), s.adj = (s.correct+1)/(s.attempt+1),
               df = s.attempt + l.attempt - 1, difference=s.adj-l.adj, t = qt(0.975, df),
               upper = difference + t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt),
               lower = difference - t*sqrt(s.adj*(1-s.adj)/s.attempt+l.adj*(1-l.adj)/l.attempt))
ggplot(data=prop) + geom_errorbar(aes(x=r, ymax=upper, ymin=lower)) + geom_point(aes(x=r, y=difference)) +
        geom_hline(aes(yint=0, colour="red")) + labs(x="Correlation", y="Scatterplot power minus Overlaid Lines") + facet_grid(n~smoothed)
@
 \caption{The difference in estimated power (scatter plot power minus overlaid line power) to reveal the true plot by smoothness and sample size with a 95\% confidence interval. The red line indicates the point where both plot designs are equally powerful while values above this line indicate that the scatter plot is more powerful than the overlaid line graph. The largest differences appear in negative correlation, where the scatter plot is more powerful.}
 \label{powerdiff}
\end{figure}
\fi

@@

Figure~\ref{samplesize} demonstrates the effects of changing sample size for unsmoothed data. The scatter plot has increased power to detect correlation almost universally for any increase. Sample size influences the overlaid lines differently, they appears to benefit for positively correlated true plots, but it does not improve the weak performance in negative correlation.
Figure~\ref{time} and figure~\ref{confidence} give a breakdown of the interactions between the response variables: plot detection, time and confidence. Figure~\ref{time} is the histogram of the time taken to record a response, split by plot design and sign of correlation and true plot detection. The time data was highly positively skewed with a maximum of 1760 seconds so the variable was log-transformed. We found that the subject could identify the true plot out of the overlaid line graph 33.6\% of the time and it took on average $e^{3.442}$ = 31.2 seconds to make a choice. Successful detections were slightly faster than unsuccessful decision, taking on average 28.7 seconds to pick the true plot and 32.6 seconds if a null was selected. The scatter plots performed much better, with the true plot picked 52.9\% of the time, with an average of 22.7 seconds required to make a decision. If the true plot was detected it took only 18.6 seconds, and 29.3 seconds for a null plot selection.

@@

There are two major reasons why a subject may fail to detect the true plot out of a lineup. Either the plot design was unable to display correlation for that particular combination of factors (and the analyst would commit a Type II error by not rejecting the null hypothesis), or a null plot that had a greater correlation than the true plot was selected instead (and the analyst would commit a Type I error by rejecting the null hypothesis). We next can analyse these kinds of errors with the pin plots found in Figure~\ref{pins}. The frequency of selection of each plot in each lineup, either true or null, is represented by a pin, placed at the value of the correlation between the two variables shown in that plot. The correlation of the selected plot is on the x-axis and the frequency of selection is plotted on the y axis. The correlation of the simulated (true) data ($\rho \in \pm \{0.3, 0.5, 0.7, 0.9\}$ in the plot, the sample size ($t \in \{12, 24, 48, 96\}$), and the plot design (unsmoothed/smoothed, scatterplot/lineplot) is used to facet the plots. The blue pins indicate the true plots and the red pins indicate the null plots. If the plot design has the power to display correlation, we would expect the highest correlation plot to be selected, leading to a Type I error where a plot from the null distribution spuriously had high correlation. We find that this is generally not the case. The selections appear to have a wide range of correlations, indicating that if a subject selected a null plot it was typically not due to excess spurious correlation. Instead the failure to select the highest correlated plot indicates that a Type II error was committed rather than Type I as the plot designs did not have the power to differentiate between medium and low correlation plots. However, for the scatter plot, whenever there was a plot with a very strong correlation, usually in the form of the true plot, it was overwhelmingly selected. For the overlaid lines this mostly occurred only for plots with strong positive correlation, which is consistent with its general lack of power to reveal negative relationships in data.

\begin{figure}[htbp]
  \centering
<<pins, dependson='setup', fig.width=6, fig.height=10, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
count.setup <- select(full, pic_id, response_0, actual_location, design, smoothed, n)
count.setup$response_0 <- as.character(count.setup$response_0)
counts <- data.frame(table(count.setup$pic_id, count.setup$response_0))
counts$Var1 <- as.numeric(as.character(counts$Var1))
counts$Var2 <- as.numeric(as.character(counts$Var2))
colnames(counts) <- c("lineup", "plot", "count")
pin <- merge(counts, bylineup, by.x="lineup", by.y="pic_id")
pin <- select(pin, lineup, plot, count, smoothed, actual_location, n, design, r, evals)
pin <- mutate(pin, r = factor(r),
              true = ifelse(plot==actual_location, "True plot", "Null plot"), 
              percent = count/evals,
              base = 0,
              correlation = diag(as.matrix(correl[pin$lineup, (pin$plot+2)])))
pin.a <- ggplot(data=filter(pin, smoothed==0 & design == "Scatterplot")) + 
         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
         geom_point(aes(x=correlation, y=percent, colour=true)) + coord_fixed() + facet_grid(n~r) + 
         labs(title ="Unsmoothed Scatterplot", y="Percent Chosen") + theme(legend.position="none")
pin.b <- ggplot(data=subset(pin, smoothed==1 & design == "Scatterplot")) + 
         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
         geom_point(aes(x=correlation, y=percent, colour=true)) + coord_fixed()  + facet_grid(n~r) + 
         labs(title="Smoothed Scatterplot", y="Percent Chosen") + theme(legend.position="none")
pin.c <- ggplot(data=subset(pin, smoothed==0 & design == "Overlaid Lines")) +
         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
         geom_point(aes(x=correlation, y=percent, colour=true)) + coord_fixed()  + facet_grid(n~r) +
         labs(title="Unsmoothed Overlaid Lines", y="Percent Chosen") + theme(legend.position="none")
pin.d <- ggplot(data=subset(pin, smoothed==1 & design == "Overlaid Lines")) + 
         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
         geom_point(aes(x=correlation, y=percent, colour=true)) + coord_fixed()  + facet_grid(n~r) + 
         labs(title="Smoothed Overlaid Lines", y="Percent Chosen") + theme(legend.position="none")
grid.arrange(pin.a, pin.b, pin.c, pin.d, ncol=1)
@
  \caption{Pin plots of the subject selections of each lineup, by design, smoothness, true plot correlaition and sample size. On the x-axis is the correlation of the plot (there typically is spurious correlation in the null plots), and on the y-axis is the proportion of times a plot was chosen. Each plot contains data from the three lineups with the requisite factor combination. Red pins indicate null plots and blue indicates the true plot, occasionally the true plot does not seem to appear as it has a proportion of zero and is visually blocked by the red pins. Strongly correlated plot seem to be selected regardless of if the relationship is real or spurious.}
  \label{pins}
\end{figure}

@@

\subsection{Modelling Graphical Power}

@@

\begin{table}[htp]
\caption{Weighted Mixed Effects Logit Model for the predicted power of plot design. Sample size for scatter plot evaluations: 900, overlaid lines: 949, total: 1849. Default plot design is overlaid lines. The model is fitted using the GLIMMIX procedure in SAS.}
\label{tab1}
\centering
<<powerfit, dependson='setup', results='asis', echo=FALSE, message=FALSE, error=FALSE, warning=FALSE>>=
powercoef <- powercoef[,-1]
rownames(powercoef) <- c("Intercept", "abs(r)", "sign(r)", "abs(r)*sign(r)", "smoothed", "t", "design", "design*abs(r)", "design*sign(r)", "design*abs(r)*sign(r)", "design*smoothed", "design*t")
colnames(powercoef) <- c("Estimate", "Standard Error", "df", "t-value", "p-value")
kable(powercoef, format="latex", digits=2)
@
\end{table}

Table \ref{tab1} summarises the model estimates. It can be confusing the interpret the model estimates, and at first glance it appears the negative plot coefficient implies that overlaid lines are substantially more powerful than the scatter plot, but this is misleading and interpreting the full model involves combining the estimates for the different interactions with $\rho$. $\beta_{2}$ and $\beta_{8}$ model the power difference between low and high correlations, while $\beta_{3}, \beta_{4}, \beta_{9}$ and $\beta_{10}$ model any increase in power appearing only in positive correlation. 
Taking these into account, the scatter plot shows superior performance relative to the overlaid lines at higher values of abs($\rho$). It needs to be restated that we only tested correlations at  $0.3, 0.5, 0.7$ and $0.9$, so the results below an absolute value of $0.3$ are unreliable and may be artifacts of regression smoothing methods.

%The correlation:sign interaction shows that the scatter plot has close to symmetric performance for positive (correlation:sign + correlation) and negative (-correlation) correlation while the overlaid line graph is substantially weaker for negative values of correlation.
Smoothing out the excess 'noise' from a time series had a minimal effect on scatter plots, it did have a significant positive impact on the power of overlaid line graphs as smoothing induces a large visual difference (Figure~\ref{smoothing}. Increasing the sample size, $t$, improved the power of the scatter plot but it was not statistically significant for the overlaid line graph.
